---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
keep_md: yes
---


## Loading and preprocessing the data
##Part 1 of assignment

```{r preparing R,echo=TRUE}
library(dplyr)
library(ggplot2)
library(knitr)

```                      


Read input file into a dataframe ignoring NAs and handling factors

```{r read data,echo=TRUE}

data <- read.csv("activity.csv", stringsAsFactors=T)
complete<-na.omit(data)
```



## What is mean total number of steps taken per day?

Perpare the sum, mean, and median

```{r group data, echo=TRUE}
daily<-group_by(complete,date) 
(per_day   <- summarise(daily, Sum= sum(steps),Mean=mean(steps), Median=median(steps)))
```


## What is the average daily activity pattern?

Draw the histogram for the sum

```{r histogram plot, echo=TRUE}
hist(per_day$Sum)
```

##Part 2 of assignment

Prepare the data to get the average for the 5 min interval
```{r}
interval<-group_by(complete,interval) 
(grouped_interval   <- summarise(interval, Average= mean(steps)))
```

Plot the average

```{r, echo=TRUE}
plot(grouped_interval$Average~grouped_interval$interval,type="l",ylab="Average Steps", xlab=" 5-minute interval",main="Average Interval Steps")
```

Finding the interval with maximum average

```{r,echo=TRUE}
Max_Int<-max(grouped_interval$Average)
grouped_interval[ which(grouped_interval$Average==Max_Int),1]
```

## Imputing missing values
##Part 3 of assignment

Compute the number of NA rows
```{r, echo=TRUE}
sub<-subset(data,is.na(data$steps)==TRUE)
dim(sub)[1]
```

Filling the missing data using the mean for the interval
Make a copy of original data so as not to loose it !!
```{r,echo=TRUE}
data1<-data

interval2<-group_by(data1,interval) 
(grouped_interval2   <- summarise(interval2, 
                                  Average= mean(steps,na.rm=TRUE)))
index<-which(is.na(data1)==TRUE)
for(i in 1:length(index))
        {
        inter1<-data1[index[i],3]
        new_steps<-grouped_interval2[which(grouped_interval2[1]
                                           == inter1),2]                                  
        data1[index[i],1]=ceiling(new_steps)
        }
## display the first 20 rows of the new dataset without NAs

head(data1,20)
```

Calcualte the new sum, mean, and median of the new dataset and draw the histogram

Perpare the sum, mean, and median

```{r}
daily2<-group_by(data1,date) 
(per_day2   <- summarise(daily2, Sum= sum(steps),Mean=mean(steps), Median=median(steps)))
```

Draw the histogram for the sum

```{r, echo=TRUE}
hist(per_day2$Sum,xlab=" Total number of steps taken each day", main="Total No. of Steps")
```

compare results before and after adding missing data
```{r before and after, echo=TRUE}
#par(mfrow=c(2,1))

v1<-as.numeric(per_day$Mean)
v2<-as.numeric(per_day$Median)
v3<-as.numeric(per_day2$Mean)
v4<-as.numeric(per_day2$Median)

plot_colors <- c("black","red")

plot(v1, type="l", col=plot_colors[1], xaxt = "n",ylab="",xlab="")
lines(v3, type="l", col=plot_colors[2])

title(ylab= "Average Steps")
title(xlab="Day")
axis(1,at=1:53,labels=per_day$date)

legend("topleft",legend=(c("Mean Without NAs","Mean with NAs")), 
       col=plot_colors,pch = c(1,2), lty = c(1,2)) 
# gives the legend lines the correct color and width


plot(v2, type="l", col=plot_colors[1], xaxt = "n",ylab="",xlab="")
lines(v4, type="l", col=plot_colors[2])

title(ylab= "Median step value")
title(xlab="Day ")
axis(1,at=1:53,labels=per_day$date)

legend("bottomleft",legend=(c("Median Without NAs","Median with NAs")),col=plot_colors,pch = c(1,2), lty = c(1,2)) 
# gives the legend lines the correct color and width
```
Effect of adding missing values on total number of daily steps
```{r effect on sum,echo=TRUE}
v5<-as.numeric(per_day$Sum)
v6<-as.numeric(per_day2$Sum)

plot_colors <- c("black","red")

plot(v5, type="l", col=plot_colors[1], xaxt = "n",ylab="",xlab="")
lines(v6, type="l", col=plot_colors[2])

title(ylab= "Total Steps")
title(xlab="Day")
axis(1,at=1:53,labels=per_day$date)

legend("topleft",legend=(c("Sum Without NAs","Sum with NAs")), col=plot_colors,pch = c(1,2), lty = c(1,2)) # gives the legend lines the correct color and width
```



## Are there differences in activity patterns between weekdays and weekends?
```{r weekday, echo=TRUE}
library(timeDate)
library(dplyr)

wend<-NULL
for(i in 1:dim(data1)[1])
        {
        wend[i]<-isWeekend(data1[i,2])
        }
data2<-cbind(data1,wend)

inter3<-subset(data2,data2$wend=="FALSE")
group1<-group_by(inter3,interval) 
(weekday_avg  <- summarise(group1,Mean=mean(steps)))
inter4<-subset(data2,data2$wend=="TRUE")
group2<-group_by(inter4,interval) 
(weekend_avg  <- summarise(group2,Mean=mean(steps)))


# plot_colors <- c("black","red")
# 
# plot(weekday_avg$Mean, type="l", col=plot_colors[1], xaxt = "n",ylab="",xlab="")
# lines(weekend_avg$Mean, type="l", col=plot_colors[2])
# 
# title(ylab= "Average Steps")
# title(xlab="Interval")
# axis(1,at=1:288,labels=weekday_avg$interval)
# 
# legend("topleft",legend=(c("Weekday Average","Weekend Average")), col=plot_colors,pch = c(1,2), lty = c(1,2))

library(ggplot2)
require(gridExtra)


p1<-ggplot(data=weekend_avg, aes(x=weekend_avg$interval, y=weekend_avg$Mean, group=1)) + 
  geom_line(size=1) + 
  xlab("5-minute interval") +
  ylab("Average Steps") +
  ggtitle("Weekday Mean Steps")

p2<-ggplot(data=weekday_avg, aes(x=weekday_avg$interval, y=weekday_avg$Mean, group=1)) + 
  geom_line(size=1) + 
    xlab("5-minute interval") +
  ylab("Average Steps") +
  ggtitle("Weekend Mean Steps")
grid.arrange(p1, p2, nrow=2)
```
